---
title: "EDA: Fertilizer treatments and winter wheat biomass"
subtitle: "Data structure, treatment rankings, and year-to-year variability"
author: "Jelena Čuklina"
format:
  pptx:
    slide-level: 2
execute:
  echo: false
  warning: false
  message: false
---

## Project question

**Main EDA question**

What are properties of ZOFE Winter wheat dataset, that are necessary to highlight for interpretation of other method outputs?

**Role of this notebook**

This presentation complements notebooks with the causal DAG/SEM and predictive regression modeling with Random Forest and Linear models by showing the empirical structure.

---

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(forcats)
library(knitr)
library(scales)
library(RColorBrewer)

resolve_repo_path <- function(...) {
  candidates <- c(".", "..")

  for (base in candidates) {
    candidate <- file.path(base, ...)
    if (file.exists(candidate) || dir.exists(candidate)) {
      return(candidate)
    }
  }

  file.path(candidates[[1]], ...)
}

ww_data_path <- resolve_repo_path("data", "processed", "zofe_ww.csv")
prep_script_path <- resolve_repo_path("src", "ZOFE_MVStat2_data_preparation.R")

if (!file.exists(ww_data_path)) {
  source(prep_script_path)
}

ZOFE_WW <- read_csv(ww_data_path, show_col_types = FALSE)

theme_set(theme_bw())

format_corr_label <- function(data, x_var, y_var) {
  corr_value <- suppressWarnings(cor(data[[x_var]], data[[y_var]], use = "pairwise.complete.obs"))
  paste0("r = ", round(corr_value, 2))
}
```

---

## Treatment-to-color mapping

Centralized treatment colors are defined here so they can be modified once and reused across all plots.

```{r}
# Qualitative categorical palette from RColorBrewer.
# Set3 is designed for categorical group distinctions and works well in presentations.
observed_treatments <- sort(unique(ZOFE_WW$treatment_en))

n_treatments <- length(observed_treatments)

if (n_treatments <= 12) {
  palette_values <- brewer.pal(max(3, n_treatments), "Set3")[1:n_treatments]
} else {
  palette_values <- colorRampPalette(
    brewer.pal(12, "Set3")
  )(n_treatments)
}

treatment_colors <- setNames(
  palette_values,
  observed_treatments
)

year_values <- sort(unique(ZOFE_WW$year))
year_shapes <- setNames(
  rep(c(0:14, 16:25), length.out = length(year_values)),
  year_values
)
```

---

## Dataset structure

```{r}
data_structure <- tibble(
  Quantity = c("Rows", "Columns", "Years", "Treatments", "Replicates", "Plots"),
  Value = c(
    nrow(ZOFE_WW),
    ncol(ZOFE_WW),
    n_distinct(ZOFE_WW$year),
    n_distinct(ZOFE_WW$treatment_en),
    n_distinct(ZOFE_WW$replicate),
    n_distinct(ZOFE_WW$plot_nr)
  )
)

kable(data_structure)
```

**Why this matters:** treatment comparisons should be interpreted together with year structure and missingness.

---

## Missingness in key variables

```{r}
#| fig-width: 9
#| fig-height: 5
vars_key <- c(
  "fert_Nmin_tot",
  "fert_N_tot",
  "fert_P_tot",
  "fert_K_tot",
  "soil_0_20_Ntot",
  "soil_0_20_P_test",
  "soil_0_20_K_test",
  "annual_N_uptake",
  "annual_P_uptake",
  "annual_K_uptake",
  "annual_total_biomass_maincrop_DM"
)

missing_summary <- ZOFE_WW %>%
  select(all_of(vars_key)) %>%
  summarise(across(
    everything(),
    ~ mean(is.na(.x)) * 100,
    .names = "{.col}"
  )) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "pct_missing"
  ) %>%
  arrange(desc(pct_missing))

missing_summary %>%
  ggplot(aes(x = fct_reorder(variable, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Missingness in key variables",
    x = NULL,
    y = "Missing values (%)"
  ) +
  theme_bw()
```

---

## Biomass distribution

```{r}
#| fig-width: 9
#| fig-height: 5
ZOFE_WW %>%
  ggplot(aes(x = annual_total_biomass_maincrop_DM)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of winter wheat biomass",
    x = "Annual total biomass of main crop (DM)",
    y = "Number of observations"
  ) +
  theme_bw()
```

**Interpretation:** this shows the total productivity range before comparing fertilizer treatments.

---

## Fertilization treatments ranked by biomass

```{r}
#| fig-width: 9
#| fig-height: 5
treatment_biomass_summary <- ZOFE_WW %>%
  group_by(treatment_en) %>%
  summarise(
    n = n(),
    mean_biomass = mean(annual_total_biomass_maincrop_DM, na.rm = TRUE),
    median_biomass = median(annual_total_biomass_maincrop_DM, na.rm = TRUE),
    sd_biomass = sd(annual_total_biomass_maincrop_DM, na.rm = TRUE),
    se_biomass = sd_biomass / sqrt(sum(!is.na(annual_total_biomass_maincrop_DM))),
    lower_mean = mean_biomass - 1.96 * se_biomass,
    upper_mean = mean_biomass + 1.96 * se_biomass,
    .groups = "drop"
  ) %>%
  arrange(desc(mean_biomass))

treatment_biomass_order <- treatment_biomass_summary$treatment_en

ZOFE_WW <- ZOFE_WW %>%
  mutate(
    treatment_en = factor(treatment_en, levels = treatment_biomass_order)
  )

treatment_colors <- treatment_colors[treatment_biomass_order]

treatment_biomass_summary %>%
  ggplot(aes(
    x = factor(treatment_en, levels = rev(treatment_biomass_order)),
    y = mean_biomass,
    fill = treatment_en
  )) +
  geom_col() +
  geom_errorbar(aes(ymin = lower_mean, ymax = upper_mean), width = 0.2) +
  scale_fill_manual(values = treatment_colors, guide = "none") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Treatment ranking by mean biomass",
    subtitle = "Error bars show approximate 95% intervals around treatment means",
    x = "Treatment",
    y = "Mean biomass"
  )
```

---

## Are treatment differences large relative to variability?

```{r}
#| fig-width: 9
#| fig-height: 5
ZOFE_WW %>%
  mutate(
    treatment_en = fct_reorder(
      treatment_en,
      annual_total_biomass_maincrop_DM,
      .fun = median,
      na.rm = TRUE
    )
  ) %>%
  ggplot(aes(
    x = treatment_en,
    y = annual_total_biomass_maincrop_DM,
    fill = treatment_en
  )) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.25) +
  scale_fill_manual(values = treatment_colors, guide = "none") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Biomass variation within and between treatments",
    x = "Treatment",
    y = "Annual total biomass of main crop (DM)"
  )
```

**Interpretation:** overlapping distributions imply that average rankings should be interpreted cautiously.

---

## Does biomass fluctuate year-to-year?

```{r}
#| fig-width: 10
#| fig-height: 5

ZOFE_WW %>%
  ggplot(aes(
    x = factor(year),
    y = annual_total_biomass_maincrop_DM,
    color = treatment_en
  )) +
  geom_boxplot(
    outlier.shape = NA,
    color = "black"
  ) +
  geom_jitter(
    width = 0.18,
    #alpha = 0.45,
    size = 1.3
  ) +
  scale_color_manual(values = treatment_colors) +
  labs(
    title = "Biomass distributions across years",
    subtitle = "Individual observations are colored by fertilization treatment",
    x = "Year",
    y = "Annual total biomass of main crop (DM)",
    color = "Treatment"
  ) +
  theme_bw()
```

---

## Year-to-year biomass anomalies

```{r}
#| fig-width: 10
#| fig-height: 5
overall_mean_biomass <- mean(
  ZOFE_WW$annual_total_biomass_maincrop_DM,
  na.rm = TRUE
)

year_anomaly <- ZOFE_WW %>%
  group_by(year) %>%
  summarise(
    mean_biomass = mean(annual_total_biomass_maincrop_DM, na.rm = TRUE),
    anomaly = mean_biomass - overall_mean_biomass,
    .groups = "drop"
  )

year_anomaly %>%
  ggplot(aes(x = year, y = anomaly)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Year-to-year biomass anomalies",
    subtitle = "Positive values indicate years above the overall dataset mean",
    x = "Year",
    y = "Biomass anomaly"
  ) +
  theme_bw()
```

**Interpretation:** large anomalies indicate that year effects may dominate simple treatment averages.

---

## Treatment trajectories across years

```{r}
#| fig-width: 10
#| fig-height: 5
treatment_year_summary <- ZOFE_WW %>%
  group_by(year, treatment_en) %>%
  summarise(
    n = n(),
    mean_biomass = mean(annual_total_biomass_maincrop_DM, na.rm = TRUE),
    .groups = "drop"
  )

treatment_year_summary %>%
  ggplot(aes(
    x = year,
    y = mean_biomass,
    group = treatment_en,
    color = treatment_en
  )) +
  geom_line(alpha = 0.75) +
  geom_point(size = 1.5) +
  scale_color_manual(values = treatment_colors) +
  labs(
    title = "Treatment-specific biomass trajectories",
    x = "Year",
    y = "Mean biomass",
    color = "Treatment"
  ) +
  theme_bw()
```

---

## Treatment ranking stability across years

```{r}
#| fig-width: 10
#| fig-height: 5
treatment_year_ranks <- treatment_year_summary %>%
  group_by(year) %>%
  mutate(
    biomass_rank = rank(-mean_biomass, ties.method = "average")
  ) %>%
  ungroup()

treatment_year_ranks %>%
  ggplot(aes(
    x = year,
    y = biomass_rank,
    group = treatment_en,
    color = treatment_en
  )) +
  geom_line(alpha = 0.75) +
  geom_point(size = 1.5) +
  scale_y_reverse(breaks = pretty_breaks()) +
  scale_color_manual(values = treatment_colors) +
  labs(
    title = "Treatment rank by biomass across years",
    subtitle = "Rank 1 means highest mean biomass in that year",
    x = "Year",
    y = "Biomass rank",
    color = "Treatment"
  ) +
  theme_bw()
```

---

## Fertilizer nutrient content

```{r}
#| fig-width: 10
#| fig-height: 5

fert_summary <- ZOFE_WW %>%
  group_by(treatment_en) %>%
  summarise(
    mean_N = mean(fert_N_tot, na.rm = TRUE),
    mean_P = mean(fert_P_tot, na.rm = TRUE),
    mean_K = mean(fert_K_tot, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "nutrient",
    values_to = "amount"
  ) %>%
  mutate(
    treatment_en = factor(treatment_en, levels = treatment_biomass_order),
    nutrient = recode(
      nutrient,
      mean_N = "Nitrogen (N)",
      mean_P = "Phosphorus (P)",
      mean_K = "Potassium (K)"
    )
  )

ggplot(
  fert_summary,
  aes(
    x = treatment_en,
    y = amount,
    fill = treatment_en
  )
) +
  geom_col() +
  facet_wrap(~ nutrient, scales = "free_y") +
  scale_fill_manual(values = treatment_colors, guide = "none") +
  labs(
    title = "Average fertilizer nutrient inputs by treatment",
    subtitle = "Averaged across all years",
    x = "Treatment",
    y = "Average nutrient input"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


--- 

## Heatmap: treatment performance by year

```{r}
#| fig-width: 10
#| fig-height: 5
treatment_year_ranks %>%
  ggplot(aes(
    x = year,
    y = fct_reorder(treatment_en, mean_biomass, .fun = mean),
    fill = mean_biomass
  )) +
  geom_tile() +
  labs(
    title = "Mean biomass by treatment and year",
    x = "Year",
    y = "Treatment",
    fill = "Mean biomass"
  ) +
  theme_bw()
```

**Interpretation:** horizontal patterns show consistently strong or weak treatments; vertical patterns show high- or low-productivity years.

---

## Fertilizer inputs and biomass

```{r}
#| fig-width: 10
#| fig-height: 5
fert_vars <- c("fert_Nmin_tot", "fert_N_tot", "fert_P_tot", "fert_K_tot")

ZOFE_WW %>%
  select(all_of(c(fert_vars, "annual_total_biomass_maincrop_DM", "treatment_en", "year"))) %>%
  pivot_longer(
    cols = all_of(fert_vars),
    names_to = "fertilizer_variable",
    values_to = "fertilizer_amount"
  ) %>%
  ggplot(aes(
    x = fertilizer_amount,
    y = annual_total_biomass_maincrop_DM,
    color = treatment_en,
    shape = factor(year)
  )) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "black", alpha = 0.45) +
  facet_wrap(~ fertilizer_variable, scales = "free_x") +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = year_shapes) +
  labs(
    title = "Fertilizer inputs and biomass",
    subtitle = "Points are colored by fertilization treatment; LOESS curves are descriptive and do not imply causality",
    x = "Fertilizer input",
    y = "Biomass",
    color = "Treatment",
    shape = "Year"
  ) +
  theme_bw()
```

---

## Nutrient uptake and biomass

```{r}
#| fig-width: 10
#| fig-height: 5
uptake_vars <- c("annual_N_uptake", "annual_P_uptake", "annual_K_uptake")

ZOFE_WW %>%
  select(all_of(c(uptake_vars, "annual_total_biomass_maincrop_DM", "treatment_en", "year"))) %>%
  pivot_longer(
    cols = all_of(uptake_vars),
    names_to = "uptake_variable",
    values_to = "uptake"
  ) %>%
  ggplot(aes(
    x = uptake,
    y = annual_total_biomass_maincrop_DM,
    color = treatment_en,
    shape = factor(year)
  )) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "black", alpha = 0.45) +
  facet_wrap(~ uptake_variable, scales = "free_x") +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = year_shapes) +
  labs(
    title = "Nutrient uptake and biomass",
    subtitle = "Descriptive context for SEM biomass-path coefficients",
    x = "Nutrient uptake",
    y = "Biomass",
    color = "Treatment",
    shape = "Year"
  ) +
  theme_bw()
```

---

## Nutrient uptake relationship

```{r}
#| fig-width: 10
#| fig-height: 5

uptake_pair_data <- bind_rows(
  ZOFE_WW %>%
    transmute(
      pair = "Nitrogen vs phosphorus",
      x_value = annual_N_uptake,
      y_value = annual_P_uptake,
      x_label = "Annual N uptake",
      y_label = "Annual P uptake",
      treatment_en,
      year
    ),
  ZOFE_WW %>%
    transmute(
      pair = "Nitrogen vs potassium",
      x_value = annual_N_uptake,
      y_value = annual_K_uptake,
      x_label = "Annual N uptake",
      y_label = "Annual K uptake",
      treatment_en,
      year
    ),
  ZOFE_WW %>%
    transmute(
      pair = "Phosphorus vs potassium",
      x_value = annual_P_uptake,
      y_value = annual_K_uptake,
      x_label = "Annual P uptake",
      y_label = "Annual K uptake",
      treatment_en,
      year
    )
)

uptake_pair_labels <- uptake_pair_data %>%
  group_by(pair) %>%
  summarise(
    x = min(x_value, na.rm = TRUE),
    y = max(y_value, na.rm = TRUE),
    label = paste0("r = ", round(cor(x_value, y_value, use = "pairwise.complete.obs"), 2)),
    .groups = "drop"
  )

ggplot(
  uptake_pair_data,
  aes(
    x = x_value,
    y = y_value,
    color = treatment_en,
    shape = factor(year)
  )
) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE, color = "black", alpha = 0.45) +
  geom_text(
    data = uptake_pair_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 4
  ) +
  facet_wrap(~ pair, scales = "free") +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = year_shapes) +
  labs(
    title = "Nutrient uptake relationship",
    subtitle = "Each panel shows one pair of uptake variables with linear trend and Pearson correlation",
    x = "First nutrient uptake variable",
    y = "Second nutrient uptake variable",
    color = "Treatment",
    shape = "Year"
  ) +
  theme_bw()
```

---

## Soil vs fertilizer nutrient scatter plots

```{r}
#| fig-width: 10
#| fig-height: 5

soil_fert_data <- bind_rows(
  ZOFE_WW %>% transmute(nutrient = "Nitrogen (N)", fertilizer_input = fert_N_tot, soil = soil_0_20_Ntot, treatment_en, year),
  ZOFE_WW %>% transmute(nutrient = "Phosphorus (P)", fertilizer_input = fert_P_tot, soil = soil_0_20_P_test, treatment_en, year),
  ZOFE_WW %>% transmute(nutrient = "Potassium (K)", fertilizer_input = fert_K_tot, soil = soil_0_20_K_test, treatment_en, year)
)

ggplot(
  soil_fert_data,
  aes(
    x = fertilizer_input,
    y = soil,
    color = treatment_en,
    shape = factor(year)
  )
) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, alpha = 0.45, color = 'grey', fill = 'grey90') +
  geom_point() +
  facet_wrap(~ nutrient, scales = "free") +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = year_shapes) +
  labs(
    title = "Soil vs fertilizer nutrient scatter plots",
    subtitle = "Points are colored by treatment and shaped by year",
    x = "Fertilizer nutrient input",
    y = "Soil nutrient measurement",
    color = "Treatment",
    shape = "Year"
  ) +
  theme_bw()
```

---

## Fertilizer vs plant uptake scatter plots

```{r}
#| fig-width: 10
#| fig-height: 5

plant_fert_uptake_data <- bind_rows(
  ZOFE_WW %>% transmute(nutrient = "Nitrogen (N)", fertilizer = fert_N_tot, uptake = annual_N_uptake, treatment_en, year),
  ZOFE_WW %>% transmute(nutrient = "Phosphorus (P)", fertilizer = fert_P_tot, uptake = annual_P_uptake, treatment_en, year),
  ZOFE_WW %>% transmute(nutrient = "Potassium (K)", fertilizer = fert_K_tot, uptake = annual_K_uptake, treatment_en, year)
)

ggplot(
  plant_fert_uptake_data,
  aes(
    x = fertilizer,
    y = uptake,
    color = treatment_en,
    shape = factor(year)
  )
) +  
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, alpha = 0.45, color = 'grey', fill = 'grey90') +
  geom_point() +
  facet_wrap(~ nutrient, scales = "free") +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = year_shapes) +
  labs(
    title = "Fertilizer vs plant uptake scatter plots",
    subtitle = "Points are colored by treatment and shaped by year",
    x = "Fertilizer nutrient input",
    y = "Plant nutrient uptake",
    color = "Treatment",
    shape = "Year"
  ) +
  theme_bw()
```

---

## Soil vs plant uptake scatter plots

```{r}
#| fig-width: 10
#| fig-height: 5

soil_uptake_data <- bind_rows(
  ZOFE_WW %>% transmute(nutrient = "Nitrogen (N)", soil = soil_0_20_Ntot, uptake = annual_N_uptake, treatment_en, year),
  ZOFE_WW %>% transmute(nutrient = "Phosphorus (P)", soil = soil_0_20_P_test, uptake = annual_P_uptake, treatment_en, year),
  ZOFE_WW %>% transmute(nutrient = "Potassium (K)", soil = soil_0_20_K_test, uptake = annual_K_uptake, treatment_en, year)
)

ggplot(
  soil_uptake_data,
  aes(
    x = soil,
    y = uptake,
    color = treatment_en,
    shape = factor(year)
  )
) +  
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, alpha = 0.45, color = 'grey', fill = 'grey90') +
  geom_point() +
  facet_wrap(~ nutrient, scales = "free") +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = year_shapes) +
  labs(
    title = "Soil vs plant uptake scatter plots",
    subtitle = "Points are colored by treatment and shaped by year",
    x = "Soil nutrient measurement",
    y = "Plant nutrient uptake",
    color = "Treatment",
    shape = "Year"
  ) +
  theme_bw()
```

---

## Correlation analysis: key numeric variables

```{r}
#| fig-width: 10
#| fig-height: 6

corr_vars <- c(
  "fert_Nmin_tot",
  "fert_N_tot",
  "fert_P_tot",
  "fert_K_tot",
  "soil_0_20_Ntot",
  "soil_0_20_P_test",
  "soil_0_20_K_test",
  "annual_N_uptake",
  "annual_P_uptake",
  "annual_K_uptake",
  "annual_total_biomass_maincrop_DM"
)

corr_data <- ZOFE_WW %>%
  select(all_of(corr_vars))

corr_matrix <- cor(
  corr_data,
  use = "pairwise.complete.obs",
  method = "pearson"
)

corr_long <- as.data.frame(as.table(corr_matrix)) %>%
  rename(
    variable_1 = Var1,
    variable_2 = Var2,
    correlation = Freq
  )

ggplot(corr_long, aes(x = variable_1, y = variable_2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_equal() +
  labs(
    title = "Correlation heatmap of fertilizer, soil, uptake, and biomass variables",
    subtitle = "Pearson correlations using pairwise-complete observations",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Interpretation:** this plot shows which variables move together before fitting the DAG/SEM. Strong correlations can support expected pathways, but correlation alone does not imply causation.

---

## Correlation structure grouped by nutrient pathway

```{r}
#| fig-width: 10
#| fig-height: 7

# Variables are manually ordered to group N, P, and K pathways together.
ordered_corr_vars <- c(
  "fert_Nmin_tot",
  "fert_N_tot",
  "soil_0_20_Ntot",
  "annual_N_uptake",
  "fert_P_tot",
  "soil_0_20_P_test",
  "annual_P_uptake",
  "fert_K_tot",
  "soil_0_20_K_test",
  "annual_K_uptake",
  "annual_total_biomass_maincrop_DM"
)

ordered_corr_data <- ZOFE_WW %>%
  select(all_of(ordered_corr_vars))

ordered_corr_matrix <- cor(
  ordered_corr_data,
  use = "pairwise.complete.obs",
  method = "pearson"
)

ordered_corr_long <- as.data.frame(as.table(ordered_corr_matrix)) %>%
  rename(
    variable_1 = Var1,
    variable_2 = Var2,
    correlation = Freq
  )

ordered_corr_long$variable_1 <- factor(
  ordered_corr_long$variable_1,
  levels = ordered_corr_vars
)

ordered_corr_long$variable_2 <- factor(
  ordered_corr_long$variable_2,
  levels = rev(ordered_corr_vars)
)

ggplot(
  ordered_corr_long,
  aes(x = variable_1, y = variable_2, fill = correlation)
) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation structure grouped by nutrient pathway",
    subtitle = "Variables are ordered to reveal within-pathway and cross-pathway relationships",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Interpretation:** this ordering makes it easier to visually assess whether nutrient pathways are internally coherent. For example, if nitrogen-related variables correlate strongly with each other, the N pathway may behave more consistently than pathways with weak internal correlations. Cross-pathway correlations may also suggest shared growth processes or nutrient interactions.

---

## Which variables correlate most with biomass?

```{r}
biomass_correlations <- corr_long %>%
  filter(
    variable_1 == "annual_total_biomass_maincrop_DM",
    variable_2 != "annual_total_biomass_maincrop_DM"
  ) %>%
  mutate(abs_correlation = abs(correlation)) %>%
  arrange(desc(abs_correlation))

kable(
  biomass_correlations %>%
    select(variable_2, correlation) %>%
    rename(
      Variable = variable_2,
      `Correlation with biomass` = correlation
    ),
  digits = 3
)
```

```{r}
#| fig-width: 9
#| fig-height: 5

biomass_correlations %>%
  ggplot(aes(
    x = fct_reorder(variable_2, correlation),
    y = correlation
  )) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Variables ranked by correlation with biomass",
    subtitle = "Positive values indicate higher biomass when the variable is higher",
    x = NULL,
    y = "Pearson correlation with biomass"
  ) +
  theme_bw()
```

**Interpretation:** this ranking provides a descriptive comparison with the SEM biomass equation. Variables with strong correlations with biomass are candidates for important productivity-related pathways, but causal interpretation still requires the DAG/SEM assumptions.

---

## Correlations among nutrient uptake variables

```{r}
#| fig-width: 8
#| fig-height: 5

uptake_corr_data <- ZOFE_WW %>%
  select(annual_N_uptake, annual_P_uptake, annual_K_uptake)

uptake_corr_matrix <- cor(
  uptake_corr_data,
  use = "pairwise.complete.obs",
  method = "pearson"
)

uptake_corr_long <- as.data.frame(as.table(uptake_corr_matrix)) %>%
  rename(
    variable_1 = Var1,
    variable_2 = Var2,
    correlation = Freq
  )

ggplot(uptake_corr_long, aes(x = variable_1, y = variable_2, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 5) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_equal() +
  labs(
    title = "Correlations among nutrient uptake variables",
    subtitle = "High correlations suggest that uptake pathways are not fully independent",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_bw()
```

**Interpretation:** if N, P, and K uptake are strongly correlated, then separate nutrient effects in SEM may partly reflect shared crop growth processes rather than fully independent nutrient pathways.

---

## Main EDA takeaways

- Fertilization treatments differ by produced biomass.
- Biomass appears to fluctuate substantially across years.
- Treatment performance should be interpreted together with year-to-year variability.
- There are noticeable issues with the data:
  - missingness in nitrogen
  - correlation of annual plant uptake variables
  - non-linear effects
  - negative correlation of soil nitrogen with other nutrient flux covariates and biomass

---

## Link to other notebooks

The EDA supports the causal analysis by showing:

1. how much biomass varies across treatments;
2. whether year effects are large;
3. whether treatment rankings are stable;
4. whether simple fertilizer, soil, and uptake relationships appear approximately linear.

The causal notebook then evaluates whether these observed patterns are consistent with a structured nutrient-pathway model. The prediction notebook then evaluates whether observed patterns might explain differences in prediction by different models
