---
title: "Zurich weather covariates from covariate dictionary"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

This Quarto document reads `weather_covariate_dictionary_updated.xlsx` and calculates the listed weather covariates for **Zurich only**.

The main function calculates covariates for **one specific calendar month only**. A second wrapper function then calculates a full season by calling the monthly function month by month and selecting only the covariates listed for each month in the Excel dictionary.

For a `target_year`, the season is defined as:

- October--December from `target_year - 1`
- January--June from `target_year`

For example, the **2022 season** is October--December 2021 plus January--June 2022.

## Important GDD note

If `gdd_base5` returned `NA` before, the likely issue was a mismatch between the dictionary formula and the prepared daily weather table:

- the dictionary formula uses `daily_tmax` and `daily_tmin`
- the prepared daily table used derived names such as `daily_max_temp_c` and `daily_min_temp_c`

This version keeps the descriptive names **and** adds aliases `daily_tmax` and `daily_tmin` before calculating GDD, so the dictionary formula can be followed closely.

```{r}
#| label: setup

# Required packages:
# install.packages(c(
#   "readxl", "dplyr", "tidyr", "lubridate", "purrr",
#   "stringr", "slider", "jsonlite", "httr2", "tibble", "readr"
# ))

library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(purrr)
library(stringr)
library(slider)
library(jsonlite)
library(httr2)
library(tibble)
library(readr)

resolve_repo_path <- function(...) {
  candidates <- c(".", "..")

  for (base in candidates) {
    candidate <- file.path(base, ...)
    if (file.exists(candidate) || dir.exists(candidate)) {
      return(candidate)
    }
  }

  file.path(candidates[[1]], ...)
}

dictionary_path <- resolve_repo_path("data", "weather_covariate_dictionary_updated.xlsx")
nutrient_flux_path <- resolve_repo_path(
  "data", "raw", "Copy of ZOFE_Mulitivariat_course_WW_2022_nutrient_fluxes.xlsx"
)
processed_dir <- resolve_repo_path("data", "processed")
dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: function-monthly-weather-covariates

max_consecutive_true <- function(x) {
  x <- tidyr::replace_na(as.logical(x), FALSE)
  if (!any(x)) return(0L)
  r <- rle(x)
  max(r$lengths[r$values])
}

read_weather_dictionary <- function(dictionary_path = "weather_covariate_dictionary_updated.xlsx") {
  month_lookup <- tibble(
    month = month.abb,
    month_num = seq_along(month.abb)
  )

  readxl::read_excel(dictionary_path) |>
    filter(!if_all(everything(), is.na)) |>
    mutate(
      month = str_to_title(as.character(month)),
      month = str_sub(month, 1, 3)
    ) |>
    left_join(month_lookup, by = "month") |>
    filter(!is.na(month_num))
}

fetch_open_meteo_hourly <- function(latitude,
                                    longitude,
                                    start_date,
                                    end_date,
                                    timezone = "Europe/Zurich") {
  # Open-Meteo Historical Weather API.
  # Dictionary variable mapping:
  # - hourly_precip_mm := Open-Meteo hourly precipitation, mm
  # - hourly_temp_c    := Open-Meteo hourly temperature_2m, deg C
  request("https://archive-api.open-meteo.com/v1/archive") |>
    req_url_query(
      latitude = latitude,
      longitude = longitude,
      start_date = as.character(start_date),
      end_date = as.character(end_date),
      hourly = "temperature_2m,precipitation",
      timezone = timezone,
      temperature_unit = "celsius",
      precipitation_unit = "mm"
    ) |>
    req_perform() |>
    resp_body_json(simplifyVector = TRUE)
}

prepare_hourly_daily_weather <- function(api_response, timezone = "Europe/Zurich") {
  hourly <- tibble(
    datetime = ymd_hm(api_response$hourly$time, tz = timezone),
    date = as.Date(datetime),
    hourly_temp_c = as.numeric(api_response$hourly$temperature_2m),
    hourly_precip_mm = as.numeric(api_response$hourly$precipitation)
  )

  daily <- hourly |>
    group_by(date) |>
    summarise(
      # Derived from hourly data because Open-Meteo is requested hourly here.
      daily_precip_mm = sum(hourly_precip_mm, na.rm = TRUE),
      daily_min_temp_c = min(hourly_temp_c, na.rm = TRUE),
      daily_max_temp_c = max(hourly_temp_c, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      # Aliases used by the Excel dictionary formula for GDD.
      # This fixes the common NA/missing-object issue when formula text uses
      # daily_tmax/daily_tmin but the prepared table uses longer variable names.
      daily_tmin = daily_min_temp_c,
      daily_tmax = daily_max_temp_c
    )

  list(hourly = hourly, daily = daily)
}

calculate_one_covariate <- function(variable_name,
                                    calculation_from_hourly_data,
                                    hourly_m,
                                    daily_m) {
  # This function follows calculation_from_hourly_data as closely as possible.
  # It uses explicit branches for safety rather than eval(parse()).
  formula_txt <- str_squish(as.character(calculation_from_hourly_data))

  if (variable_name == "rain_total_mm" || formula_txt == "sum(hourly_precip_mm)") {
    return(sum(hourly_m$hourly_precip_mm, na.rm = TRUE))
  }

  if (variable_name == "consecutive_dry_days" || str_detect(formula_txt, "daily_precip_mm < 1")) {
    return(max_consecutive_true(daily_m$daily_precip_mm < 1))
  }

  if (variable_name == "heavy_rain_days_20mm" || str_detect(formula_txt, "daily_precip_mm > 20")) {
    return(sum(daily_m$daily_precip_mm > 20, na.rm = TRUE))
  }

  if (variable_name == "max_3day_rain_mm" || str_detect(formula_txt, "rollsum\\(daily_precip_mm, 3\\)")) {
    x <- daily_m$daily_precip_mm
    if (length(x) < 3) return(NA_real_)
    rolling_3day <- slider::slide_dbl(x, sum, .before = 2, .complete = TRUE, na.rm = TRUE)
    return(max(rolling_3day, na.rm = TRUE))
  }

  if (variable_name == "temp_mean_c" || formula_txt == "mean(hourly_temp_c)") {
    return(mean(hourly_m$hourly_temp_c, na.rm = TRUE))
  }

  if (variable_name == "heat_days_gt28" || str_detect(formula_txt, "daily_max_temp_c > 28")) {
    return(sum(daily_m$daily_max_temp_c > 28, na.rm = TRUE))
  }

  if (variable_name == "heat_days_gt30" || str_detect(formula_txt, "daily_max_temp_c > 30")) {
    return(sum(daily_m$daily_max_temp_c > 30, na.rm = TRUE))
  }

  if (variable_name == "gdd_base5" || str_detect(formula_txt, "daily_tmax") || str_detect(formula_txt, "daily_tmin")) {
    # Dictionary: sum(pmax(((daily_tmax + daily_tmin)/2) - 5, 0))
    # Open-Meteo does not directly return daily_tmax/daily_tmin in this script;
    # they are derived from hourly temperature_2m and stored as aliases above.
    return(sum(
      pmax(((daily_m$daily_tmax + daily_m$daily_tmin) / 2) - 5, 0),
      na.rm = TRUE
    ))
  }

  if (variable_name == "frost_days_lt0" || str_detect(formula_txt, "daily_min_temp_c < 0")) {
    return(sum(daily_m$daily_min_temp_c < 0, na.rm = TRUE))
  }

  if (variable_name == "diurnal_temp_range" || str_detect(formula_txt, "daily_max_temp_c - daily_min_temp_c")) {
    return(mean(daily_m$daily_max_temp_c - daily_m$daily_min_temp_c, na.rm = TRUE))
  }

  warning("No calculation rule implemented for variable_name = ", variable_name)
  NA_real_
}

extract_zurich_weather_covariates_one_month <- function(calendar_year,
                                                        month_num,
                                                        dictionary_path = "weather_covariate_dictionary_updated.xlsx",
                                                        latitude = 47.3769,
                                                        longitude = 8.5417,
                                                        timezone = "Europe/Zurich") {
  stopifnot(length(calendar_year) == 1, length(month_num) == 1)
  stopifnot(month_num >= 1, month_num <= 12)

  start_date <- as.Date(sprintf("%d-%02d-01", calendar_year, month_num))
  end_date <- start_date + lubridate::days(lubridate::days_in_month(start_date) - 1L)
  days_in_selected_month <- as.integer(lubridate::days_in_month(start_date))

  dict <- read_weather_dictionary(dictionary_path) |>
    filter(month_num == !!month_num)

  if (nrow(dict) == 0) {
    warning("No dictionary covariates listed for ", month.abb[month_num], " ", calendar_year)
    return(tibble())
  }

  raw <- fetch_open_meteo_hourly(
    latitude = latitude,
    longitude = longitude,
    start_date = start_date,
    end_date = end_date,
    timezone = timezone
  )

  weather <- prepare_hourly_daily_weather(raw, timezone = timezone)

  hourly_m <- weather$hourly |>
    filter(date >= start_date, date <= end_date)

  daily_m <- weather$daily |>
    filter(date >= start_date, date <= end_date)

  if (nrow(hourly_m) == 0 || nrow(daily_m) == 0) {
    warning("No weather data returned for ", month.abb[month_num], " ", calendar_year)
  }

  dict |>
    mutate(
      location = "Zurich",
      latitude = latitude,
      longitude = longitude,
      calendar_year = calendar_year,
      month_num = month_num,
      start_date = start_date,
      end_date = end_date,
      days_in_month = days_in_selected_month,
      observed_days = n_distinct(daily_m$date),
      observed_hours = nrow(hourly_m),
      value = pmap_dbl(
        list(variable_name, calculation_from_hourly_data),
        \(v, f) calculate_one_covariate(v, f, hourly_m, daily_m)
      ),
      covariate_id = paste(variable_name, tolower(month), calendar_year, sep = "_")
    ) |>
    select(
      location,
      latitude,
      longitude,
      calendar_year,
      month,
      month_num,
      start_date,
      end_date,
      days_in_month,
      observed_days,
      observed_hours,
      variable_name,
      covariate_id,
      group,
      unit,
      aggregation_level,
      definition,
      calculation_from_hourly_data,
      value,
      priority,
      main_responses
    ) |>
    arrange(group, variable_name)
}
```

```{r}
#| label: function-period-weather-covariates

season_month_calendar <- function(target_year,
                                  months_previous_year = 10:12,
                                  months_target_year = 1:6) {
  bind_rows(
    tibble(calendar_year = target_year - 1L, month_num = months_previous_year),
    tibble(calendar_year = target_year, month_num = months_target_year)
  ) |>
    distinct(calendar_year, month_num) |>
    arrange(calendar_year, month_num)
}

extract_zurich_weather_covariates_period <- function(target_year,
                                                     dictionary_path = "weather_covariate_dictionary_updated.xlsx",
                                                     latitude = 47.3769,
                                                     longitude = 8.5417,
                                                     timezone = "Europe/Zurich",
                                                     months_previous_year = 10:12,
                                                     months_target_year = 1:6) {
  months_to_calculate <- season_month_calendar(
    target_year = target_year,
    months_previous_year = months_previous_year,
    months_target_year = months_target_year
  )

  covariates_long <- pmap_dfr(
    months_to_calculate,
    \(calendar_year, month_num) {
      extract_zurich_weather_covariates_one_month(
        calendar_year = calendar_year,
        month_num = month_num,
        dictionary_path = dictionary_path,
        latitude = latitude,
        longitude = longitude,
        timezone = timezone
      )
    }
  ) |>
    mutate(season_target_year = target_year, .before = calendar_year) |>
    arrange(calendar_year, month_num, group, variable_name)

  # For the season-level wide table, keep the season target year as its own
  # metadata column, but do not include the calendar year in each covariate
  # column name. Example: rain_total_mm_oct instead of
  # rain_total_mm_oct_2021. This is safe for one Oct--Jun season because each
  # month appears only once within a season.
  covariates_wide <- covariates_long |>
    mutate(covariate_id_wide = paste(variable_name, tolower(month), sep = "_")) |>
    select(location, season_target_year, covariate_id_wide, value) |>
    pivot_wider(names_from = covariate_id_wide, values_from = value)

  list(
    covariates_long = covariates_long,
    covariates_wide = covariates_wide,
    months_calculated = months_to_calculate,
    source = "Open-Meteo Historical Weather API: https://open-meteo.com/en/docs/historical-weather-api"
  )
}
```

```{r}
#| label: monthly-examples

# Single-month checks. These are useful for validating that the function is
# filtering weather data to exactly one calendar month.

zurich_october_2021 <- extract_zurich_weather_covariates_one_month(
  calendar_year = 2021,
  month_num = 10,
  dictionary_path = dictionary_path
)

zurich_february_2020 <- extract_zurich_weather_covariates_one_month(
  calendar_year = 2020,
  month_num = 2,
  dictionary_path = dictionary_path
)

zurich_february_2022 <- extract_zurich_weather_covariates_one_month(
  calendar_year = 2022,
  month_num = 2,
  dictionary_path = dictionary_path
)

zurich_april_2022 <- extract_zurich_weather_covariates_one_month(
  calendar_year = 2022,
  month_num = 4,
  dictionary_path = dictionary_path
)

zurich_october_2021
zurich_february_2020
zurich_february_2022
zurich_april_2022

readr::write_csv(
  zurich_october_2021,
  file.path(processed_dir, "zurich_weather_covariates_october_2021.csv")
)
readr::write_csv(
  zurich_february_2020,
  file.path(processed_dir, "zurich_weather_covariates_february_2020.csv")
)
readr::write_csv(
  zurich_february_2022,
  file.path(processed_dir, "zurich_weather_covariates_february_2022.csv")
)
readr::write_csv(
  zurich_april_2022,
  file.path(processed_dir, "zurich_weather_covariates_april_2022.csv")
)
```

```{r}
#| label: zurich-2022-season

# The 2022 season starts in October 2021 and continues through June 2022.
# The period function calls the one-month function month by month and, for each
# month, selects only the covariates listed for that month in the Excel dictionary.

zurich_2022_season <- extract_zurich_weather_covariates_period(
  target_year = 2022,
  dictionary_path = dictionary_path,
  months_previous_year = 10:12,
  months_target_year = 1:6
)

zurich_2022_season$covariates_long
zurich_2022_season$covariates_wide

# Save both long and wide forms.
readr::write_csv(
  zurich_2022_season$covariates_long,
  file.path(processed_dir, "zurich_weather_covariates_2022_season_long.csv")
)

readr::write_csv(
  zurich_2022_season$covariates_wide,
  file.path(processed_dir, "zurich_weather_covariates_2022_season_wide.csv")
)

# GDD diagnostic: should no longer be NA unless source data for April is missing.
zurich_2022_season$covariates_long |>
  filter(variable_name == "gdd_base5") |>
  select(calendar_year, month, days_in_month, observed_days, observed_hours,
         variable_name, calculation_from_hourly_data, value)
```

```{r}
#| label: zurich-years-1995-2022
years <- 1995:2022
years_weather <- list()
for (year in years) {
  this_year <- extract_zurich_weather_covariates_period(
    target_year = year,
    dictionary_path = dictionary_path,
    months_previous_year = 10:12,
    months_target_year = 1:6
  )
  years_weather[[as.character(year)]] <- this_year$covariates_wide
}
df_years <- do.call(rbind, years_weather)
```
```{r}
readr::write_csv(
  df_years,
  file.path(processed_dir, "zurich_weather_covariates_1995_2022_seasons.csv")
)
```

```{r}
#| label: merge-weather-with-nutrient-flux

library(readxl)
nutrient_flux_data <- read_excel("data/raw/ZOFE_Mulitivariat_course_WW_2022_nutrient_fluxes.xlsx")
nutrient_flux_data <- read_excel(nutrient_flux_path, 
    skip = 1)

nutrient_flux_data <- readxl::read_excel(
  nutrient_flux_path,
  col_names = FALSE
) |>
  setNames(as.character(unlist(.[1, ]))) |>
  slice(-c(1, 2)) |>
  mutate(year = as.integer(year))

weather_with_nutrient_flux <- df_years |>
  left_join(
    nutrient_flux_data,
    by = join_by(season_target_year == year)
  )

weather_with_nutrient_flux

readr::write_csv(
  weather_with_nutrient_flux,
  file.path(processed_dir, "zurich_weather_covariates_1995_2022_seasons_merged.csv")
)
```
