---
title: "02 — Fertilizer similarity and data limitations"
subtitle: "Comparing fertilizer effects on grain yield under year/weather variation and small-sample uncertainty"
author: "Project team"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

# Purpose of this notebook

This notebook investigates whether different fertilizers have distinguishable effects on final grain yield.

The central scientific question is:

> **Do different fertilizers produce similar or distinguishable effects on grain yield?**

This is deliberately framed as a **similarity and uncertainty** question, not only as a hypothesis test for differences.

Because the dataset contains only a limited number of observations, the goal is not to produce a definitive fertilizer ranking. Instead, this notebook asks:

1. Are fertilizer differences visible in the raw data?
2. Are fertilizer differences stable across years?
3. Do estimated fertilizer effects remain after adjusting for year and weather?
4. Which fertilizers appear similar within the uncertainty of the data?
5. Where does the dataset reach its inferential limit?

# Relationship to other notebooks

This notebook builds directly on Notebook 1.

Notebook 1 identified:

- yield variation across fertilizers,
- year-to-year variability,
- possible weather effects,
- multicollinearity among nutrient variables,
- small-sample limitations.

This notebook focuses specifically on fertilizer-level differences in grain yield.

Outputs from this notebook will inform:

- Notebook 3: whether fertilizer identity should be included as a predictor in yield models;
- Notebook 4: how fertilizer effects should be interpreted causally and mechanistically.

# Setup

```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(broom)
library(emmeans)
library(performance)
library(patchwork)
library(lme4)
library(lmerTest)
```

# Load data

```{r}
# Load cleaned dataset from Notebook 1

dat <- readRDS("outputs/cleaned_data.rds")

# Inspect structure
str(dat)
```

# Basic preprocessing

```{r}
dat <- dat %>%
  mutate(
    fertilizer_type = as.factor(fertilizer_type),
    year = as.factor(year)
  )
```

# Variable groups

```{r}
weather_vars <- c(
  "rain_total_mm_mar",
  "rain_total_mm_apr",
  "consecutive_dry_days_may",
  "heat_days_gt30_may",
  "heat_days_gt30_jun",
  "gdd_base5_apr"
)

response_var <- "grain_yield"
```

# Exploratory fertilizer comparison

## Observations per fertilizer and year

Before fitting models, inspect whether the design is balanced.

```{r}
dat %>%
  count(year, fertilizer_type) %>%
  arrange(year, fertilizer_type)
```

```{r}
dat %>%
  count(fertilizer_type) %>%
  ggplot(aes(x = fertilizer_type, y = n)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Number of observations per fertilizer",
    x = "Fertilizer",
    y = "Number of observations"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Raw yield by fertilizer

```{r}
ggplot(dat, aes(x = fertilizer_type, y = grain_yield, color = fertilizer_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2) +
  theme_minimal() +
  labs(
    title = "Raw grain yield by fertilizer",
    subtitle = "Boxplots with individual replicate observations",
    x = "Fertilizer",
    y = "Grain yield"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

## Mean yield and uncertainty by fertilizer

```{r}
fert_summary <- dat %>%
  group_by(fertilizer_type) %>%
  summarise(
    n = n(),
    mean_yield = mean(grain_yield, na.rm = TRUE),
    sd_yield = sd(grain_yield, na.rm = TRUE),
    se_yield = sd_yield / sqrt(n),
    ci_low = mean_yield - qt(0.975, df = n - 1) * se_yield,
    ci_high = mean_yield + qt(0.975, df = n - 1) * se_yield,
    .groups = "drop"
  )

fert_summary
```

```{r}
ggplot(fert_summary, aes(x = reorder(fertilizer_type, mean_yield), y = mean_yield)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Mean grain yield by fertilizer",
    subtitle = "Error bars show approximate 95% confidence intervals",
    x = "Fertilizer",
    y = "Mean grain yield"
  )
```

## Draft interpretation

Replace after analysis.

> The raw fertilizer comparison showed [clear / moderate / weak] separation between fertilizers. Several fertilizers had strongly overlapping yield distributions, suggesting that the data may not support a precise fertilizer ranking. This motivates model-based comparisons that adjust for year and weather while keeping uncertainty central.

# Year dependence of fertilizer effects

## Yield by year and fertilizer

```{r}
ggplot(dat, aes(x = fertilizer_type, y = grain_yield, color = fertilizer_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7) +
  facet_wrap(~ year) +
  theme_minimal() +
  labs(
    title = "Grain yield by fertilizer within each year",
    x = "Fertilizer",
    y = "Grain yield"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

## Fertilizer × year mean trends

```{r}
fert_year_summary <- dat %>%
  group_by(year, fertilizer_type) %>%
  summarise(
    mean_yield = mean(grain_yield, na.rm = TRUE),
    .groups = "drop"
  )

 ggplot(fert_year_summary, aes(x = year, y = mean_yield, group = fertilizer_type, color = fertilizer_type)) +
  geom_line(alpha = 0.8) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Fertilizer-specific mean yield across years",
    subtitle = "Crossing lines suggest fertilizer × year interaction or unstable ranking",
    x = "Year",
    y = "Mean grain yield"
  )
```

## Draft interpretation

> The year-wise plots showed that fertilizer ranking was [stable / unstable / partly stable] across years. If fertilizer lines cross strongly, this suggests that fertilizer effects may depend on environmental conditions or that the data are too noisy to rank fertilizers reliably.

# Model 1: fertilizer-only model

## Scientific question

> Are there average yield differences between fertilizers, ignoring year and weather?

This is a baseline descriptive model.

```{r}
mod_fert_only <- lm(grain_yield ~ fertilizer_type, data = dat)
summary(mod_fert_only)
anova(mod_fert_only)
confint(mod_fert_only)
```

## Model diagnostics

```{r}
performance::check_model(mod_fert_only)
```

## Estimated marginal means

```{r}
emm_fert_only <- emmeans(mod_fert_only, ~ fertilizer_type)
emm_fert_only
```

```{r}
plot(emm_fert_only) +
  labs(
    title = "Estimated fertilizer means: fertilizer-only model",
    x = "Estimated grain yield",
    y = "Fertilizer"
  )
```

## Pairwise contrasts

```{r}
contrast_fert_only <- pairs(emm_fert_only, adjust = "tukey")
contrast_fert_only
```

```{r}
contrast_fert_only_df <- broom::tidy(contrast_fert_only)
contrast_fert_only_df
```

## Draft interpretation

> In the fertilizer-only model, fertilizer type explained [little / moderate / substantial] variation in grain yield. However, because this model ignores year and weather, it should be interpreted as a descriptive comparison rather than an adjusted fertilizer effect estimate.

# Model 2: year-adjusted fertilizer model

## Scientific question

> Do fertilizer differences remain after accounting for year-to-year variation?

Year adjustment is important because the dataset contains observations from five years, and weather/environmental differences may strongly affect yield.

```{r}
mod_fert_year <- lm(grain_yield ~ fertilizer_type + year, data = dat)
summary(mod_fert_year)
anova(mod_fert_year)
confint(mod_fert_year)
```

## Model diagnostics

```{r}
performance::check_model(mod_fert_year)
```

## Estimated fertilizer means adjusted for year

```{r}
emm_fert_year <- emmeans(mod_fert_year, ~ fertilizer_type)
emm_fert_year
```

```{r}
plot(emm_fert_year) +
  labs(
    title = "Estimated fertilizer means adjusted for year",
    x = "Estimated grain yield",
    y = "Fertilizer"
  )
``